Large Language Models in Reinforcement Learning - A Comparison of Methods

### 2023-11-13 doc/paper.tex
Update paper with peer review changes
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

The agent is trained as in the baseline. The agent's state observation is encoded to natural text, and sent as part of a prompt to the LLM at every timestep. After being given the state observation, the LLM is asked to respond with a list of recommended actions for the agent to perform. The agent then acts independently of the LLM for that timestep, as dictated by the baseline RL algorithm it is trained with. As it does so, its chosen action is encoded to natural text. Then, another LLM is used for semantic similarity comparison, comparing the proposed action to the action performed by the agent. Finally, high semantic similarity gives an additional pseudo-reward which is added to the reward of the environment. Compared to the policy influencing method above, this approach is less direct, as the policy is only indirectly affected by the LLM through these additional rewards, which only take effect when the policy is updated through backpropogation after an entire episode is completed. Thus, this method functions more as an advisor, influencing the policy in a more long term capacity than the coaching of the policy influencer. Throughout the paper, we will refer to this as LLM reward shaping.

@@

Looking at LLM policy influencing, we see that it has resulted in somewhat worse results than using the PPO baseline. Averaging per episode over all eight runs shows that LLM policy influencing only achieves $93\%$ of the reward amount per episode that the baseline achieves. Judging from the graph, it may seem that policy influencing may have given some initial learning advantage, but ultimately slows learning for the agent further on. As with reward shaping, we have also used caching of LLM responses to significantly decrease the computational overhead cost to a minimum.

@@

As we can see, LLM reward shaping performs the best overall, and performs much better than vanilla PPO, which struggles to learn the correct policy in this complex environment. Given the terrible performance of the baseline, LLM reward shaping achieved a percentage increase of $275\%$, which is not that impressive when the average reward from the baseline was close to zero. The baseline achieved an average reward of $0.0097$, while LLM reward shaping achieved an average reward of $0.037$. See Table \ref{unlockpickuptable}.

@@

As shown in the previously referenced work (\ref{background}), the integration of a Large Language Model into the Reinforcement Learning framework can improve network performance. There are various ways to do this, and our work attempts to compare two main methodologies of integrating the LLM as well as the combination of the two, namely through reward shaping and policy influencing. We have implemented algorithms to perform this on top of a PPO baseline, and have measured several interesting results across environments of varying complexities.

>>>>

In this method, the agent is trained using the same method as in the baseline. The agent's state observation is encoded to natural text, and sent as part of a prompt to the LLM at every timestep. After being given the state observation, the LLM is asked to respond with a list of recommended actions for the agent to perform. The agent then acts independently of the LLM for that timestep, as dictated by the baseline RL algorithm it is trained with. As it does so, its chosen action is encoded to natural text. Then, another LLM is used for semantic similarity comparison, comparing the proposed action to the action performed by the agent. Finally, high semantic similarity gives an additional pseudo-reward which is added to the reward of the environment. Compared to the policy influencing method above, this approach is less direct, as the policy is only indirectly affected by the LLM through these additional rewards, which only take effect when the policy is updated through backpropogation after an entire episode is completed. Thus, this method functions more as an advisor, influencing the policy in a more long term capacity than the coaching of the policy influencer. Throughout the paper, we will refer to this as LLM reward shaping.

@@

Looking at LLM policy influencing, we see that it has resulted in somewhat worse results than using the PPO baseline. Averaging over all eight runs shows that LLM policy influencing only achieves $93\%$ of the reward amount the baseline receives. Judging from the graph, it may seem that policy influencing may have given some initial learning advantage, but ultimately slows learning for the agent further on. As with reward shaping, we have also used caching of LLM responses to significantly decrease the computational overhead cost to a minimum.

@@

As we can see, LLM reward shaping performs the best overall, and performs much better than the baseline PPO, which struggles to learn the correct policy in this complex environment. Given the terrible performance of the baseline, LLM reward shaping achieved a percentage increase of $275\%$, which is not that impressive when the average reward from the baseline was close to zero. The baseline achieved an average reward of $0.0097$, while LLM reward shaping achieved an average reward of $0.037$. See Table \ref{unlockpickuptable}.

@@

As shown in the previously referenced work (Section \ref{background}), the integration of a Large Language Model into the Reinforcement Learning framework can improve network performance. There are various ways to do this, and our work attempts to compare two main methodologies of integrating the LLM as well as the combination of the two, namely through reward shaping and policy influencing. We have implemented algorithms to perform this on top of a PPO baseline, and have measured several interesting results across environments of varying complexities.
